---
title: "Incorporating Expert Range Maps in SDMs"
author: Cory Merow, Adam Wilson
output:
  html_document:
    code_folding: show
---

```{r, echo=FALSE, message=FALSE, results='hide', purl=FALSE, include=FALSE}
input  = knitr::current_input()  # filename of input document
output = paste(tools::file_path_sans_ext(input), 'R', sep = '.')
knitr::opts_chunk$set(cache=TRUE)
knitr::read_chunk('cm_knitr_settings.r')
```

<div>
<object data="3_5_assets/Minxent_Short.pdf" type="application/pdf" width="100%" height="650px"> 
  <p>It appears you don't have a PDF plugin for this browser.
   No biggie... you can <a href="3_5_assets/Minxent_Short.pdf">click here to
  download the PDF file.</a></p>  
 </object>
 </div>
 
 <p><a href="3_5_assets/Minxent_Short.pdf">Download the PDF of the presentation</a></p>  

[<i class="fa fa-file-code-o fa-3x" aria-hidden="true"></i> The R Script associated with this page is available here](`r output`).  Download this file and open it (or copy-paste into a new script) with RStudio so you can follow along.  

# Setup

## Load necessary libraries
```{r, message=F}
library(ggplot2)
library(foreach)
registerDoSEQ()
library(dplyr)
library(tidyr)
library(bossMaps)
library(raster)
```

### Load example data for the Solitary Tinamou
```{r}
species=c("Tinamus_solitarius")

data("Tinamus_solitarius_points")
data("Tinamus_solitarius_range")
data("Tinamus_solitarius_env")

points=Tinamus_solitarius_points
range=Tinamus_solitarius_range
env=Tinamus_solitarius_env
```

# Calculate Priors

## Calculate range distances
<div class="well">
Calculate the distance between expert range edge and all cells on the map (useful for estimating how likelihood of the species should decay as you get further from the expert map).

<button data-toggle="collapse" class="btn btn-primary btn-sm round" data-target="#demo1">Kind of a digression....</button>
<div id="demo1" class="collapse">

```{r, message=FALSE}
rdist = rangeDist(
  range = range,
  points = points,
  domain = env,
  domainkm = 1000,
  mask = F,
  fact = 2
  )

## Crop environmental data to this new domain (based on domainkm)
env_domain = crop(env, rdist)

## Mask pixels with no environmenta data (over ocean, etc.)
rdist = mask(rdist, env_domain[[1]])
names(rdist) = "rangeDist"
  
```
</div>
</div>

This is the result:
```{r}
plot(rdist)
plot(range, add = T)
plot(points, add = T)
```

## Evaluate curves describing expert accuracy beyond the expert map boundaries

<div class="well">
Calculate which curve parameter combinations are feasible given the domain and range geometry.

<button data-toggle="collapse" class="btn btn-primary btn-sm round" data-target="#demo2">Kind of a digression....</button>
<div id="demo2" class="collapse">

```{r,fig.width=7}
rates=checkRates(rdist)
```

From these, choose a few 
```{r}
vars=expand.grid(
  prob=c(0.5, 0.7, 0.9),
  rate=c(0,0.01,0.1,10),
  skew=0.5,
  shift=0,
  stringsAsFactors=F)
x=seq(-150,500,len=1000)
```

Calculate all the curves
```{r}
uvars=unique(vars[,c("rate","skew","shift")])
erd=do.call(rbind,
            lapply(1:nrow(uvars),function(i) {
              y=logistic(x,
                         parms=unlist(c(lower=0,upper=1,uvars[i,])))  
              return(cbind.data.frame(
                group=i,
                c(uvars[i,]),
                x=x,
                y=y))
            })
)  
```

Visualize potential decay parameters
```{r}
ggplot(erd,
       aes(
       x = x,
       y = y,
       linetype = as.factor(skew),
       colour = as.factor(rate),
       group = group
       )) +
       geom_vline(aes(xintercept = 0), colour = "red") +
       geom_line() +
       xlab("Prior value (not normalized)") +
       xlab("Distance to range edge (km)")
```
</div>
</div>

## Build priors
<div class="well">

Calculate the priors and project them across the landscape as rasters.

<button data-toggle="collapse" class="btn btn-primary btn-sm round" data-target="#demo3">Kind of a digression....</button>
<div id="demo3" class="collapse">     

Calculate frequency table of distances: In order to speed up optimization of range offsets, we calculate the `rangeOffset()` on a frequency distribution instead of the full raster.  Calculate that now with `freq()`.

```{r}
dists = freq(rdist, useNA = "no", digits = 2)
knitr::kable(head(dists))
```

```{r,warning=F, results="hide", message=F}
mcoptions <- list(preschedule = FALSE, set.seed = FALSE)
foreach(i = 1:nrow(vars), .options.multicore = mcoptions) %do% {
  ## calculate the expert range prior
  fo = paste0(species, "_prior_", paste(vars[i, ], collapse = "_"), ".grd")
  if (file.exists(fo))  return(NULL)
  expert = rangeOffset(
                      rdist,
                      parms = unlist(vars[i, ]),
                      dists = dists,
                      doNormalize = T,
                      verbose = T,
                      doWriteRaster = T,
                      filename = fo,
                      overwrite = T,
                      datatype = "FLT4S"
                      )
}
```
</div>
</div>

## Stack the priors

Here's where we'll pick up the analysis, assuming you've somehow created a reasonable prior already (unfold code above for an example with expert maps). Priors for your analyses might come from sampling bias models, maps of related species, or expert maps, among many other possibilities (Merow et al. 2016 GEB).
```{r, message=FALSE}
fs = list.files(pattern = "prior.*grd$",
                full = T,
                recursive = F)
                
priors = stack(fs)
                
## build prior table from metadata
priorf = foreach(i = 1:nlayers(priors), 
                 .combine = bind_rows) %do% {
  t1 = metadata(priors[[i]])
  t2 = t1$parms
  names(t2) = t1$pnames
  return(data.frame(id = i, t(t2)))
  }

names(priors) = paste0("prior", priorf$id)#basename(fs[wp])
plot(priors)
```

# MaxEnt/Poisson Point Process SDM

## Assemble modeling dataset
```{r}
## build single raster stack of all needed data (env and priors)
rdata = stack(env, priors)

## generate presence and non-detection datasets
pres = cbind.data.frame(
  presence = 1, 
  raster::extract(rdata, points, df =
  T, ID = F))
ns = 10000
abs = cbind.data.frame(
  presence = 0,
  ID = 1:ns,
  sampleRandom(rdata, ns))

data = rbind.data.frame(
  pres, 
  abs)
```

## Fit models
You can fit a maxent-style model or Poisson point process with `glm()` simply by applying weights to the data. See the appendix of Renner et al. 2015 (Point process models for presence-only analysis; GEB) for details on why this works. Note that I'm not fitting all the flexible functional forms of a maxent model; this is equivalent to turning off threshold and hinge features and note using regularization.

```{r, warning=F}
data$weight = 1e-6
best.var = names(env)

# number of non-NA cells
nc = cellStats(!is.na(priors[[1]]), sum)

data$weight[data$presence == 0] = nc / sum(data$presence == 0)

## Set up models
formulas = paste(
  "presence/weight ~",
  " offset(log(",
  grep("prior", 
       colnames(data),
       value = T),'))+',
  paste0(sapply(
    best.var, function(ii) {
      sapply(best.var, function(jj) {
        paste0(ii, "*", jj)
    })
  }), collapse = '+'),
  '+',
  paste0('I(', best.var, '^2)', collapse = '+'),
  '+',
  paste0(best.var, collapse = ':')
)

formulas[1]

mods = foreach(f = formulas) %do% {
  glm(
    as.formula(f),
    family = poisson(link = log),
    data = data,
    weights = weight
  )
}
```

Calculate AIC
```{r}
priorf$AIC=unlist(lapply(mods,function(x) AIC(x)))
```

## Make spatial predictions
```{r, message=FALSE}
mcoptions <- list(preschedule = FALSE, set.seed = FALSE)

ptype = "response"
psi = 1:nrow(priorf)

ps = foreach(i = psi, 
  .options.multicore = mcoptions) %dopar% {
    fo = paste0(
      species,"_posterior_",
      priorf$prob[i],"_",
      priorf$rate[i],"_",
      priorf$skew[i],"_",
      priorf$shift[i],
      ".grd"
      )
  if (file.exists(fo))
    return(NULL)
  #if(!file.exists(fo)) return("NOT YET")
  normalize(
    raster::predict(
      rdata,
      mods[[i]],
      type = ptype),
    file = fo,
    overwrite = T)
}

psf = list.files(pattern = "posterior.*.grd", full = T)
ps = stack(psf)
names(ps) = sub("prior", "posterior", names(priors)[psi])
```

## Predictions maps for different priors/offesets
<!-- fig.cap=fig$cap('preds','Predictions for different expert map smoothing parameters.', col="blue"), fig.align="center"} -->
The first column corresponds to a maxent prediction without an offset/prior (so those are are the same). Note that those models have the worst AIC compared to any of the other models, which incorporate the expert map. Figuring out which of the expert map priors is best will require some evaluation of the expert's typical accuracy, e.g. across a taxonomic group. That's the next phase of analysis...
```{r predplot,fig.width=11,fig.height=11,message=F,echo=FALSE} 
res = 1e4
p_psn = rasterVis::gplot(ps, max = res) + geom_raster(aes(fill = value))
p_psn$data$id = as.numeric(sub("prior|posterior", "", p_psn$data$variable))
p_psn$data$rate = priorf$rate[p_psn$data$id]
p_psn$data$prob = priorf$prob[p_psn$data$id]

p_psn +
  facet_grid(prob ~ rate, labeller = label_both) +
  scale_fill_gradientn(
    colours = hcols(100, bias = .5),
    trans = "log10",
    name = "Relative Occurrence Rate\np(x|Y=1)",
    na.value = "transparent"
  ) +
  geom_polygon(
    aes(x = long, y = lat, group = group),
    data = fortify(range),
    fill = "transparent",
    col = "black",
    size = .2
  ) +
  geom_point(
    aes(x = coords.x1, y = coords.x2),
    data = as.data.frame(coordinates(points[points$presence == 1,])),
    col = "black",
    size = .5
  ) +
  geom_text(
    aes(
      -30,-40,
      label = paste0("AIC=", round(AIC),
                     ifelse(
                       round(fitbuffer) == 0, "",
                       paste0("\nFitDist=", round(fitbuffer), " km")
                     )),
      group = NULL
    ),
    data = priorf,
    hjust = 1,
    size = 2
  ) +
  ylab("Latitude") + xlab("Longitude") +
  coord_equal()
```

<!-- # not working figure caption code -->
<!-- # ```{r, fig.cap=fig$cap("cos_wav", "Cosine curve", col="red"), fig.align="center"} -->
<!-- # par(mar=c(2.5, 2.5, 0.5, 0.5)); plot(x=0:100*pi/50, y=cos(0:100*pi/50), cex=0.5, xlab="x", ylab="cos(x)") -->
<!-- # ``` -->

